---
title: "Tutorial: Using daggleR"
---

The daggleR package provides two sets of helpers: in-step functions for use inside pipeline steps, and API wrappers for controlling daggle from an R session. This tutorial shows how to migrate from raw protocol calls to daggleR, and how to use the API wrappers.

## Part 1: In-step helpers

In-step helpers replace raw `cat()` and `Sys.getenv()` calls with validated, readable functions. They work inside any R step executed by daggle.

### Emitting outputs

**Before:**

```r
cat("::daggle-output name=row_count::", nrow(data), "\n")
```

**After:**

```r
daggleR::output("row_count", nrow(data))
```

`output()` validates the key format, coerces the value to character, and prints the marker. It returns the value invisibly.

### Reading upstream outputs

**Before:**

```r
n <- as.integer(Sys.getenv("DAGGLE_OUTPUT_EXTRACT_ROW_COUNT"))
```

**After:**

```r
n <- daggleR::get_output("extract", "row_count")
```

`get_output()` handles the uppercasing and hyphen-to-underscore conversion. It returns the value as a character string (you still need `as.integer()` or `as.numeric()` if you want a number).

### Run metadata

**Before:**

```r
run_id <- Sys.getenv("DAGGLE_RUN_ID")
dag <- Sys.getenv("DAGGLE_DAG_NAME")
dir <- Sys.getenv("DAGGLE_RUN_DIR")
```

**After:**

```r
run_id <- daggleR::run_id()
dag <- daggleR::dag_name()
dir <- daggleR::run_dir()
```

### Full example

A step using daggleR:

```r
library(daggleR)

raw_count <- get_output("extract", "n_raw")
cat(sprintf("Upstream extracted %s rows\n", raw_count))

data <- readRDS("data/clean.rds")
cat(sprintf("Clean data has %d rows\n", nrow(data)))

output("n_clean", nrow(data))
output("pct_retained", round(nrow(data) / as.integer(raw_count) * 100, 1))
```

In-step helpers are base R only. They require no network access and no running daggle server -- they communicate through environment variables and stdout.

### Summaries, metadata, and validations {#in-step-phase7}

Rich per-step metadata helpers, added in daggleR 0.2.0. Each emits the corresponding marker protocol documented in the [Output Protocol](../reference/output-protocol.qmd) reference.

#### `summary_md()`

Emit a markdown summary for display in dashboards:

```{r}
#| eval: false
daggleR::summary_md("## Results\n\nProcessed **42** rows successfully.")
```

Equivalent to printing `::daggle-summary format=markdown::...`.

#### `meta_numeric()`

Emit a numeric metadata entry:

```{r}
#| eval: false
daggleR::meta_numeric("row_count", 1542)
```

#### `meta_text()`

Emit a text metadata entry:

```{r}
#| eval: false
daggleR::meta_text("model_desc", "Linear regression with 3 predictors")
```

#### `meta_table()`

Emit a table metadata entry (serialized as JSON):

```{r}
#| eval: false
daggleR::meta_table("top5", head(results, 5))
```

The data frame is converted to a JSON array of objects.

#### `meta_image()`

Emit an image metadata entry (file path):

```{r}
#| eval: false
daggleR::meta_image("residuals", "output/residuals.png")
```

#### `validation()`

Emit a structured validation result:

```{r}
#| eval: false
daggleR::validation("row_count", "pass", "Expected > 0, got 1542")
daggleR::validation("schema", "fail", "Column 'date' expected date, got character")
```

The `status` argument must be one of `"pass"`, `"warn"`, or `"fail"`.

## Part 2: API wrappers

API wrappers talk to the daggle REST API (started with `daggle serve --port 8787`). They require the httr2 package.

The base URL is resolved from (in order): the `base_url` parameter, the `DAGGLE_API_URL` environment variable, or `http://127.0.0.1:8787`.

All code below assumes a running daggle server. The blocks are shown with `eval: false` since there is no live instance during doc rendering.

### List DAGs

```{r}
#| eval: false
library(daggleR)

dags <- list_dags()
dags
#>       name steps   schedule last_status            last_run
#> 1 etl-daily     4 0 3 * * *     success 2025-01-15T03:00:12Z
#> 2   reports     2       <NA>     running 2025-01-15T10:30:00Z
```

### Trigger a run and poll for completion

```{r}
#| eval: false
result <- trigger("my-pipeline", params = list(date = "2026-04-04"))
run_id <- result$run_id

repeat {
  run <- get_run("my-pipeline", run_id)
  if (run$status %in% c("success", "failure", "cancelled")) break
  Sys.sleep(2)
}

cat("Final status:", run$status, "\n")
```

The `trigger()` function returns immediately with a run ID. Poll with `get_run()` until the status is terminal.

### Get step outputs

```{r}
#| eval: false
outputs <- get_outputs("my-pipeline", run_id)
outputs
#>     step_id          key                     value
#> 1   extract    row_count                      1432
#> 2     clean      n_clean                      1398
#> 3     clean    n_dropped                        34
```

Pass `"latest"` as the run ID to get outputs from the most recent run:

```{r}
#| eval: false
outputs <- get_outputs("my-pipeline", "latest")
```

### Approve a waiting step

If a pipeline has an `approve:` step, it pauses until someone approves or rejects:

```{r}
#| eval: false
approve("my-pipeline", run_id = "latest", step_id = "review")
```

To reject instead:

```{r}
#| eval: false
reject("my-pipeline", run_id = "latest", step_id = "review")
```

### View step logs

```{r}
#| eval: false
log <- get_step_log("my-pipeline", "latest", "extract")
cat(log$stdout)
```

### Health check

```{r}
#| eval: false
health()
#> $status
#> [1] "ok"
#>
#> $version
#> [1] "0.5.0"
```

### Clean up old runs

```{r}
#| eval: false
cleanup("30d")
#> $removed
#> [1] 42
#>
#> $freed
#> [1] "10.0 MB"
```

### Artifacts, summaries, metadata, validations, and comparison {#api-phase7}

API wrappers added in daggleR 0.2.0. They talk to endpoints from daggle's Phase 7 analyst workflow features.

#### `list_artifacts()`

List declared artifacts for a run:

```{r}
#| eval: false
artifacts <- daggleR::list_artifacts("my-pipeline", "latest")
artifacts
#>   step_id   name              path   hash      size format
#> 1   train  model output/model.rds  e3b0c4   204800    rds
```

#### `plan()`

Preview the execution plan with cache status before triggering:

```{r}
#| eval: false
daggleR::plan("my-pipeline")
#>      step_id   status                      reason
#> 1      fetch outdated          input file changed
#> 2  transform   cached inputs unchanged since last run
#> 3     report no-cache         caching not enabled
```

#### `get_summaries()`

Retrieve step summaries emitted during a run:

```{r}
#| eval: false
summaries <- daggleR::get_summaries("my-pipeline", "latest")
cat(summaries$content[1])
```

#### `get_metadata()`

Retrieve typed metadata entries from a run:

```{r}
#| eval: false
meta <- daggleR::get_metadata("my-pipeline", "latest")
meta
#>   step_id       name    type  value
#> 1   train  row_count numeric   1542
#> 2   train model_desc    text Linear regression
```

#### `get_validations()`

Retrieve validation results from a run:

```{r}
#| eval: false
vals <- daggleR::get_validations("my-pipeline", "latest")
vals
#>     step_id       name status                               message
#> 1  validate  row_count   pass              Expected > 0, got 1542
#> 2  validate     schema   fail Column 'date' expected date, got char
```

#### `compare_runs()`

Compare two runs side-by-side:

```{r}
#| eval: false
diff <- daggleR::compare_runs("my-pipeline", run1 = "20250115T030012", run2 = "20250116T030015")
diff$duration_diff
#> $run1_seconds
#> [1] 47.2
#>
#> $run2_seconds
#> [1] 52.8
#>
#> $diff_seconds
#> [1] 5.6
```

### Annotations and impact (daggleR 0.3.0) {#api-phase8}

API wrappers added in daggleR 0.3.0, covering daggle's Phase 8 collaboration and observability endpoints.

#### `list_annotations()` and `add_annotation()`

Attach free-form notes to a run — useful for post-mortems (\"DB was down — manual restart at 08:30\") that should surface next to the run in status output and dashboards:

```{r}
#| eval: false
daggleR::add_annotation("penguin-monitor", "latest",
  note = "Restarted after upstream feed recovered",
  author = "alice")

daggleR::list_annotations("penguin-monitor", "latest")
#>              timestamp  author                                     note
#> 1 2026-04-17T08:45:12Z   alice  Restarted after upstream feed recovered
```

`author` defaults to `Sys.getenv("USER")`.

#### `get_impact()`

List DAGs that depend on a given DAG (via `trigger.on_dag.name`) plus any exposures declared on it (Shiny apps, Quarto reports, dashboards). Purely informational — daggle does not deploy or monitor the exposures.

```{r}
#| eval: false
impact <- daggleR::get_impact("penguin-clean")
impact$downstream_dags
#>              name project trigger_on_status
#> 1 penguin-models    data          completed
#> 2 penguin-report    data                any

impact$exposures
#>             name      type                             url description
#> 1  ops-dashboard dashboard  https://dash.example.com/ops Main ops
```

### List and filter DAGs by ownership

`list_dags()` accepts `tag`, `team`, and `owner` filters that map directly to the `?tag=&team=&owner=` query params on `GET /api/v1/dags`:

```{r}
#| eval: false
etl_dags <- daggleR::list_dags(tag = "etl", team = "data")
```

When a DAG declares `owner:`, `team:`, `description:`, or `tags:`, those fields appear as columns in the response data frame.

## When to use which

| Scenario | Use |
|---|---|
| Inside a pipeline step | In-step helpers (`output()`, `get_output()`) |
| Interactive R session | API wrappers (`trigger()`, `get_run()`) |
| Shiny app controlling pipelines | API wrappers |
| CI script kicking off a pipeline | API wrappers |
| Reading upstream data in a step | In-step `get_output()` |
| Post-mortem notes on a failed run | `add_annotation()` |
| Auditing downstream consumers before a schema change | `get_impact()` |
