---
title: "Tutorial: Using daggleR"
---

The daggleR package provides two sets of helpers: in-step functions for use inside pipeline steps, and API wrappers for controlling daggle from an R session. This tutorial shows how to migrate from raw protocol calls to daggleR, and how to use the API wrappers.

## Part 1: In-step helpers

In-step helpers replace raw `cat()` and `Sys.getenv()` calls with validated, readable functions. They work inside any R step executed by daggle.

### Emitting outputs

**Before:**

```r
cat("::daggle-output name=row_count::", nrow(data), "\n")
```

**After:**

```r
daggleR::output("row_count", nrow(data))
```

`output()` validates the key format, coerces the value to character, and prints the marker. It returns the value invisibly.

### Reading upstream outputs

**Before:**

```r
n <- as.integer(Sys.getenv("DAGGLE_OUTPUT_EXTRACT_ROW_COUNT"))
```

**After:**

```r
n <- daggleR::get_output("extract", "row_count")
```

`get_output()` handles the uppercasing and hyphen-to-underscore conversion. It returns the value as a character string (you still need `as.integer()` or `as.numeric()` if you want a number).

### Run metadata

**Before:**

```r
run_id <- Sys.getenv("DAGGLE_RUN_ID")
dag <- Sys.getenv("DAGGLE_DAG_NAME")
dir <- Sys.getenv("DAGGLE_RUN_DIR")
```

**After:**

```r
run_id <- daggleR::run_id()
dag <- daggleR::dag_name()
dir <- daggleR::run_dir()
```

### Full example

A step using daggleR:

```r
library(daggleR)

raw_count <- get_output("extract", "n_raw")
cat(sprintf("Upstream extracted %s rows\n", raw_count))

data <- readRDS("data/clean.rds")
cat(sprintf("Clean data has %d rows\n", nrow(data)))

output("n_clean", nrow(data))
output("pct_retained", round(nrow(data) / as.integer(raw_count) * 100, 1))
```

In-step helpers are base R only. They require no network access and no running daggle server -- they communicate through environment variables and stdout.

## Part 2: API wrappers

API wrappers talk to the daggle REST API (started with `daggle serve --port 8787`). They require the httr2 package.

The base URL is resolved from (in order): the `base_url` parameter, the `DAGGLE_API_URL` environment variable, or `http://127.0.0.1:8787`.

All code below assumes a running daggle server. The blocks are shown with `eval: false` since there is no live instance during doc rendering.

### List DAGs

```{r}
#| eval: false
library(daggleR)

dags <- list_dags()
dags
#>       name steps   schedule last_status            last_run
#> 1 etl-daily     4 0 3 * * *     success 2025-01-15T03:00:12Z
#> 2   reports     2       <NA>     running 2025-01-15T10:30:00Z
```

### Trigger a run and poll for completion

```{r}
#| eval: false
result <- trigger("my-pipeline", params = list(date = "2026-04-04"))
run_id <- result$run_id

repeat {
  run <- get_run("my-pipeline", run_id)
  if (run$status %in% c("success", "failure", "cancelled")) break
  Sys.sleep(2)
}

cat("Final status:", run$status, "\n")
```

The `trigger()` function returns immediately with a run ID. Poll with `get_run()` until the status is terminal.

### Get step outputs

```{r}
#| eval: false
outputs <- get_outputs("my-pipeline", run_id)
outputs
#>     step_id          key                     value
#> 1   extract    row_count                      1432
#> 2     clean      n_clean                      1398
#> 3     clean    n_dropped                        34
```

Pass `"latest"` as the run ID to get outputs from the most recent run:

```{r}
#| eval: false
outputs <- get_outputs("my-pipeline", "latest")
```

### Approve a waiting step

If a pipeline has an `approve:` step, it pauses until someone approves or rejects:

```{r}
#| eval: false
approve("my-pipeline", run_id = "latest", step_id = "review")
```

To reject instead:

```{r}
#| eval: false
reject("my-pipeline", run_id = "latest", step_id = "review")
```

### View step logs

```{r}
#| eval: false
log <- get_step_log("my-pipeline", "latest", "extract")
cat(log$stdout)
```

### Health check

```{r}
#| eval: false
health()
#> $status
#> [1] "ok"
#>
#> $version
#> [1] "0.5.0"
```

### Clean up old runs

```{r}
#| eval: false
cleanup("30d")
#> $removed
#> [1] 42
#>
#> $freed
#> [1] "10.0 MB"
```

## When to use which

| Scenario | Use |
|---|---|
| Inside a pipeline step | In-step helpers (`output()`, `get_output()`) |
| Interactive R session | API wrappers (`trigger()`, `get_run()`) |
| Shiny app controlling pipelines | API wrappers |
| CI script kicking off a pipeline | API wrappers |
| Reading upstream data in a step | In-step `get_output()` |
